C7: Fast software transactional memory for dynamic languages

### 2014-05-06 talk/dls2014/report.tex
fix some formatting

@@

% For example:
% ------------
%\address{School\\
%		 Department\\
%		 Address}
%
% Two addresses (uncomment and modify for two-address case).
% ----------------------------------------------------------
\twoauthors
 		{Remigius Meier}
 		{Department of Computer Science\\
 ETH Zürich\\
 		Switzerland}
  		{Armin Rigo}
 		{www.pypy.org}

@@

\makeatother

\usepackage{listings}
\lstset{backgroundcolor={\color{verylightgray}},
basicstyle={\scriptsize\ttfamily},
commentstyle={\ttfamily\color{commentgreen}},
keywordstyle={\bfseries\color{darkblue}},
morecomment={[l]{//}},
morekeywords={foreach,in,def,type,dynamic,Int,Boolean,infer,void,super,if,boolean,int,else,while,do,extends,class,assert,for,switch,case,private,protected,public,const,final,static,interface,new,true,false,null,return}}
\renewcommand{\lstlistingname}{Listing}

\begin{document}
%\ninept
\maketitle 
\begin{abstract}
... 
\end{abstract}

@@

STM, how atomicity \& isolation

reasons for overhead

@@

\protect\caption{Segment Addressing\label{fig:Segment-Addressing}}

\end{figure*}

@@

In order to eliminate the prohibitive memory requirements of keeping
around $N$ segment copies, we share memory between them. The segments
are initially allocated in a single range of virtual memory by a call
to \inputencoding{latin9}\lstinline!mmap()!\inputencoding{utf8}.
As illustrated in Figure \ref{fig:mmap()-Page-Mapping}, \inputencoding{latin9}\lstinline!mmap()!\inputencoding{utf8}
creates a mapping between a range of virtual memory pages and virtual
file pages. The virtual file pages are then mapped lazily by the kernel
to real physical memory pages. The mapping generated by \inputencoding{latin9}\lstinline!mmap()!\inputencoding{utf8}
is initially linear but can be changed arbitrarily. Especially, we
can remap so that multiple virtual memory pages map to a single virtual
file page. This is what we use to share memory between the segments
since then we also only require one page of physical memory.

@@

To provide atomicity for a transaction, we want to make changes visible
on commit. We also need to be able to completely abort a transaction
without a trace, like it never happened.
\begin{description}
\item [{Commit:}] If a transaction commits, we synchronize all threads
so that all of them are waiting in a safe point. In the committing
transaction, we go through all objects in the write set and check
if another transaction in a different segment read the same object.
Conflicts are resolved again by either the committing or the other
transaction waiting or aborting.\\
We then push all changes of modified objects in private pages to all
the pages in other segments, including the sharing-segment (segment
0).
\item [{Abort:}] On abort the transaction will forget about all the changes
it has done. All objects in the write set are reset by copying their
previous version from the sharing-segment into the private pages of
the aborting transaction.
\item [{}]~
\end{description}

@@

\inputencoding{latin9}\begin{lstlisting}[basicstyle={\footnotesize\ttfamily},tabsize=4]
void stm_start_transaction(tl, jmpbuf)
void stm_commit_transaction()
void stm_read(object_t *obj)
void stm_write(object_t *obj)
object_t *stm_allocate(ssize_t size_rounded)
STM_PUSH_ROOT(tl, obj)
STM_POP_ROOT(tl, obj)
\end{lstlisting}
\inputencoding{utf8}

\inputencoding{latin9}\lstinline!stm_start_transaction()!\inputencoding{utf8}
starts a transaction. It requires two arguments, the first being a
thread-local data structure and the second a buffer for use by \inputencoding{latin9}\lstinline!setjmp()!\inputencoding{utf8}.
\inputencoding{latin9}\lstinline!stm_commit_transaction()!\inputencoding{utf8}
tries to commit the current transaction. \inputencoding{latin9}\lstinline!stm_read()!\inputencoding{utf8},
\inputencoding{latin9}\lstinline!stm_write()!\inputencoding{utf8}
perform a read or a write barrier on an object and \inputencoding{latin9}\lstinline!stm_allocate()!\inputencoding{utf8}
allocates a new object with the specified size (must be a multiple
of 16). \inputencoding{latin9}\lstinline!STM_PUSH_ROOT()!\inputencoding{utf8}
and \inputencoding{latin9}\lstinline!STM_POP_ROOT()!\inputencoding{utf8}
push and pop objects on the shadow stack%
\footnote{A stack for pointers to GC objects that allows for precise garbage
collection. All objects on that stack are never seen as garbage and
are thus always kept alive.%
}. Objects have to be saved using this stack around calls that may
cause a GC cycle to happen, and also while there is no transaction
running. In this simplified API, only \inputencoding{latin9}\lstinline!stm_allocate()!\inputencoding{utf8}
and \inputencoding{latin9}\lstinline!stm_commit_transaction()!\inputencoding{utf8}
require saving object references.

The type \inputencoding{latin9}\lstinline!object_t!\inputencoding{utf8}
is special as it causes the compiler%
\footnote{Clang 3.5 with some patches to this address-space 256 feature%
} to make all accesses through it relative to the $\%gs$ register.
With exceptions, nearly all accesses to objects managed by the TM
system should use this type so that the CPU will translate the reference
to the right version of the object.

@@

On startup, we reserve a big range of virtual memory with a call to
\inputencoding{latin9}\lstinline!mmap()!\inputencoding{utf8} and
partition this space into $N+1$ segments. We want to run $N$ threads
in parallel while segment 0 is designated as the \emph{sharing-segment
}that is never assigned to a thread.

The next step involves using \inputencoding{latin9}\lstinline!remap_file_pages()!\inputencoding{utf8},
a Linux system call, to establish the fully-shared configuration.
All pages of segments $>0$ map to the pages of the sharing-segment.

However, the layout of a segment is not uniform and we actually privatize
a few areas again right away. These areas are illustrated in Figure
\ref{fig:Segment-Layout} and explained here:
\begin{description}[noitemsep]
\item [{NULL~page:}] This page is unmapped and will produce a segmentation
violation when accessed. We use this to detect erroneous dereferencing
of \inputencoding{latin9}\lstinline!NULL!\inputencoding{utf8} references.
All $\%gs{::}SO$ translated to linear addresses will point to NULL
pages if SO is set to \inputencoding{latin9}\lstinline!NULL!\inputencoding{utf8}.
\item [{Segment-local~data:}] Some area private to the segment that contains
segment-local information.
\item [{Read~markers:}] These are pages that store information about which
objects were read in the current transaction running in this segment.
\item [{Nursery:}] This area contains all the freshly allocated objects
(\emph{young objects}) of the current transaction. The GC uses pointer-bump
allocation in this area to allocate objects in the first generation.
\item [{Old~object~space:}] These pages are the ones that are really
shared between segments. They mostly contain old objects but also
some young ones that were too big to allocate in the nursery.
\end{description}
\begin{figure*}[t]
\begin{centering}
\includegraphics[scale=0.8]{\string"segment layout\string".pdf}
\par\end{centering}

@@

To improve this situation even more, we introduce the concept of \emph{overflow
objects}. If a minor collection needs to occur during a transaction,
we empty the nursery and mark each surviving object in the old object
space with an \inputencoding{latin9}\lstinline!overflow_number!\inputencoding{utf8}
globally unique to the current transaction. That way we can still
detect in a medium-fast path inside barriers that the object still
belongs to the current transaction.

@@

As seen in the API (section \ref{sub:Application-Programming-Interfac}),
we use a \emph{shadow stack }in order to provide precise garbage collection.
Any time we call a function that possibly triggers a collection, we
need to save the objects that we need afterwards on the shadow stack
using \inputencoding{latin9}\lstinline!STM_PUSH_ROOT()!\inputencoding{utf8}.
That way, they will not be freed. And in case they were young, we
get their new location in the old object space when getting them back
from the stack using \inputencoding{latin9}\lstinline!STM_POP_ROOT()!\inputencoding{utf8}.

@@

To add the object to the read set, for us it is enough to mark it
as read. Since this information needs to be local to the segment,
we need to store it in private pages. The area is called \emph{read
markers }and already mentioned in section \ref{sub:Setup}. This area
can be seen as a continuous array of bytes that is indexed from the
start of the segment by an object's reference ($SO$) divided by 16
(this requires objects of at least 16 bytes in size). Instead of just
setting the byte to \inputencoding{latin9}\lstinline!true!\inputencoding{utf8}
if the corresponding object was read, we set it to a \inputencoding{latin9}\lstinline!read_version!\inputencoding{utf8}
belonging to the transaction, which will be incremented on each commit.
Thereby, we can avoid resetting the bytes to \inputencoding{latin9}\lstinline!false!\inputencoding{utf8}
on commit and only need to do this every 255 transactions. The whole
code for the barrier is easily optimizable for compilers as well as
perfectly predictable for CPUs:

\inputencoding{latin9}\begin{lstlisting}[basicstyle={\footnotesize\ttfamily},tabsize=4]
void stm_read(SO):
    *(SO >> 4) = read_version
\end{lstlisting}
\inputencoding{utf8}

@@

The \textbf{fast path} of the write barrier is very simple. We only
need to check for the flag \inputencoding{latin9}\lstinline!WRITE_BARRIER!\inputencoding{utf8}
in the object's header and call the slow path if it is set. This flag
is set either if the object is old and comes from an earlier transaction,
or if there was a minor collection which will add the flag again on
all objects. It is never set on freshly allocated objects that still
reside in the nursery.

\inputencoding{latin9}\begin{lstlisting}[basicstyle={\footnotesize\ttfamily},tabsize=4]
void stm_write(SO):
	if SO->flags & WRITE_BARRIER:
		write_slowpath(SO)
\end{lstlisting}
\inputencoding{utf8}

@@

\inputencoding{latin9}\begin{lstlisting}[basicstyle={\footnotesize\ttfamily},tabsize=4]
void write_slowpath(SO):
	// GC part:
	list_append(to_trace, SO)
	if is_overflow_obj(SO):
		SO->flags &= ~WRITE_BARRIER
		return
	// STM part
	stm_read(SO)
	lock_idx = SO >> 4
  retry:
	if write_locks[lock_idx] == our_num:
		// we already own it
	else if write_locks[lock_idx] == 0:
		if cmp_and_swap(&write_locks[lock_idx],
					    0, our_num):
			list_append(modified_old_objects, SO)
			privatize_pages(SO)
		else:
			goto retry
	else:
		w_w_contention_management()
		goto retry
	SO->flags &= ~WRITE_BARRIER
\end{lstlisting}
\inputencoding{utf8}

First comes the \emph{GC part}: In any case, the object will be added
to the list of objects that need tracing in the next minor collection
(\inputencoding{latin9}\lstinline!to_trace!\inputencoding{utf8}).
This is necessary in case we write a reference to it that points to
a young object. We then need to trace it during the next minor collection
in order to mark the young object alive and to update its reference
to the new location it gets moved to. The check for \inputencoding{latin9}\lstinline!is_overflow_obj()!\inputencoding{utf8}
tells us if the object was actually created in this transaction. In
that case, we do not need to execute the following \emph{TM part}.
We especially do not need to privatize the page since no other transaction
knows about these ``old'' objects.

For TM, we first perform a read barrier on the object. We then try
to acquire its write lock. \inputencoding{latin9}\lstinline!write_locks!\inputencoding{utf8}
again is a simple global array of bytes that is indexed with the SO
of the object divided by 16. If we already own the lock, we are done.
If someone else owns the lock, we will do a write-write contention
management that will abort either us or the current owner of the object.
If we succeed in acquiring the lock using an atomic \inputencoding{latin9}\lstinline!cmp_and_swap!\inputencoding{utf8},
we need to add the object to the write set (a simple list called \inputencoding{latin9}\lstinline!modified_old_objects!\inputencoding{utf8})
and privatize all pages belonging to it (copy-on-write).

In all cases, we remove the \inputencoding{latin9}\lstinline!WRITE_BARRIER!\inputencoding{utf8}
flag from the object before we return. Thus, we never trigger the
slow path again before we do the next minor collection (also part
of a commit) or we start the next transaction.

@@

Aborting a transaction is rather easy. The first step is to reset
the nursery and all associated data structures. The second step is
to go over all objects in the write set (\inputencoding{latin9}\lstinline!modified_old_objects!\inputencoding{utf8})
and reset any modifications in our private pages by copying from the
sharing-segment. What is left is to use \inputencoding{latin9}\lstinline!longjmp()!\inputencoding{utf8}
to jump back to the location initialized by a \inputencoding{latin9}\lstinline!setjmp()!\inputencoding{utf8}
in \inputencoding{latin9}\lstinline!stm_start_transaction()!\inputencoding{utf8}.
Increasing the \inputencoding{latin9}\lstinline!read_version!\inputencoding{utf8}
is also done there.

@@

Committing a transaction needs a bit more work. First, we synchronize
all threads so that the committing one is the only one running and
all the others are waiting in a safe point. We then go through the
write set (\inputencoding{latin9}\lstinline!modified_old_objects!\inputencoding{utf8})
and check the corresponding \inputencoding{latin9}\lstinline!read_markers!\inputencoding{utf8}
in other threads/segments. If we detect a read-write conflict, we
do contention management to either abort us or the other transaction,
or to simply wait a bit.

After verifying that there are no conflicts anymore, we copy all our
changes done to the objects in the write set to all other segments,
including the sharing-segment. This is safe since we synchronized
all threads. We also need to push overflow objects generated by minor
collections to other segments, since they may reside partially in
private pages. At that point we also get a new \inputencoding{latin9}\lstinline!overflow_number!\inputencoding{utf8}
by increasing a global one, so that it stays globally unique for each
transaction. Increasing the \inputencoding{latin9}\lstinline!read_version!\inputencoding{utf8}
is then done at the start of a new transaction.

@@

A requirement for performing a commit is to synchronize all threads
so that we can safely update objects in other segments. To make this
synchronization fast and cheap, we do not want to insert an additional
check regularly in order to see if synchronization is requested. We
use a trick relying on the fact that dynamic languages are usually
very high-level and thus allocate a lot of objects very regularly.
This is done through the function \inputencoding{latin9}\lstinline!stm_allocate!\inputencoding{utf8}
shown below:

\inputencoding{latin9}\begin{lstlisting}[basicstyle={\footnotesize\ttfamily},tabsize=4]
object_t *stm_allocate(ssize_t size_rounded):
    result = nursery_current
	nursery_current += size_rounded
	if nursery_current > nursery_end:
		return allocate_slowpath(size_rounded)
	return result
\end{lstlisting}
\inputencoding{utf8}

This code does simple pointer-bump allocation in the nursery. If there
is still space left in the nursery, we return \inputencoding{latin9}\lstinline!nursery_current!\inputencoding{utf8}
and bump it up by \inputencoding{latin9}\lstinline!size_rounded!\inputencoding{utf8}.
The interesting part is the check \inputencoding{latin9}\lstinline!nursery_current > nursery_end!\inputencoding{utf8}
which will trigger the slow path of the function to possibly perform
a minor collection in order to free up space in the nursery.

If we want to synchronize all threads, we can rely on this check being
performed regularly. So what we do is to set the \inputencoding{latin9}\lstinline!nursery_end!\inputencoding{utf8}
to $0$ in all segments that we want to synchronize. The mentioned
check will then fail in those segments and call the slow path. In
\inputencoding{latin9}\lstinline!allocate_slowpath!\inputencoding{utf8}
they can simply check for this condition and enter a safe point.

@@

\textbf{}

>>>>

\newcommand{\mynote}[2]{%
  \textcolor{red}{%
    \fbox{\bfseries\sffamily\scriptsize#1}%
    {\small$\blacktriangleright$\textsf{\emph{#2}}$\blacktriangleleft$}%
  }%
}
\newcommand\remi[1]{\mynote{Remi}{#1}}

@@

% For example:
% ------------
%\address{School\\
%		 Department\\
%		 Address}
%
% Two addresses (uncomment and modify for two-address case).
% ----------------------------------------------------------
\twoauthors
    {Remigius Meier}
    {Department of Computer Science\\
      ETH Zürich, Switzerland\\
      \nolinkurl{remi.meier@inf.ethz.ch}}
    {Armin Rigo}
    {www.pypy.org\\
      \nolinkurl{arigo@tunes.org}}

@@

\usepackage{listings}

\lstset{backgroundcolor={\color{verylightgray}},
  basicstyle={\scriptsize\ttfamily},
  commentstyle={\ttfamily\color{commentgreen}},
  keywordstyle={\bfseries\color{darkblue}},
  morecomment={[l]{//}},
  morekeywords={foreach,in,def,type,dynamic,Int,
    Boolean,infer,void,super,if,boolean,int,else,
    while,do,extends,class,assert,for,switch,case,
    private,protected,public,const,final,static,
    interface,new,true,false,null,return}}
\renewcommand{\lstlistingname}{Listing}

\begin{document}
\ninept
\maketitle
\begin{abstract}
  ...
\end{abstract}

@@

\remi{STM, how atomicity \& isolation; reasons for overhead}

@@

    \protect\caption{Segment Addressing\label{fig:Segment-Addressing}}
\end{figure*}

@@

In order to eliminate the prohibitive memory requirements of keeping
around $N$ segment copies, we share memory between them. The segments
are initially allocated in a single range of virtual memory by a call
to \lstinline!mmap()!.  As illustrated in Figure
\ref{fig:mmap()-Page-Mapping}, \lstinline!mmap()! creates a mapping
between a range of virtual memory pages and virtual file pages. The
virtual file pages are then mapped lazily by the kernel to real
physical memory pages. The mapping generated by \lstinline!mmap()! is
initially linear but can be changed arbitrarily. Especially, we can
remap so that multiple virtual memory pages map to a single virtual
file page. This is what we use to share memory between the segments
since then we also only require one page of physical memory.

@@

To provide atomicity for a transaction, we want to make changes
visible on commit. We also need to be able to completely abort a
transaction without a trace, like it never happened.
\begin{description}
\item [{Commit:}] If a transaction commits, we synchronize all threads
  so that all of them are waiting in a safe point. In the committing
  transaction, we go through all objects in the write set and check if
  another transaction in a different segment read the same object.
  Conflicts are resolved again by either the committing or the other
  transaction waiting or aborting.\\
  We then push all changes of modified objects in private pages to all
  the pages in other segments, including the sharing-segment (segment
  0).
\item [{Abort:}] On abort the transaction will forget about all the
  changes it has done. All objects in the write set are reset by
  copying their previous version from the sharing-segment into the
  private pages of the aborting transaction.
\end{description}

@@

\begin{lstlisting}[basicstyle={\footnotesize\ttfamily},tabsize=4]
  void stm_start_transaction(tl, jmpbuf)
  void stm_commit_transaction()
  void stm_read(object_t *obj)
  void stm_write(object_t *obj)
  object_t *stm_allocate(ssize_t size_rounded)
  STM_PUSH_ROOT(tl, obj)
  STM_POP_ROOT(tl, obj)
\end{lstlisting}

\lstinline!stm_start_transaction()!  starts a transaction. It requires
two arguments, the first being a thread-local data structure and the
second a buffer for use by \lstinline!setjmp()!.
\lstinline!stm_commit_transaction()!  tries to commit the current
transaction. \lstinline!stm_read()!, \lstinline!stm_write()!  perform
a read or a write barrier on an object and \lstinline!stm_allocate()!
allocates a new object with the specified size (must be a multiple of
16). \lstinline!STM_PUSH_ROOT()!  and \lstinline!STM_POP_ROOT()!  push
and pop objects on the shadow stack~\footnote{A stack for pointers to
  GC objects that allows for precise garbage collection. All objects
  on that stack are never seen as garbage and are thus always kept
  alive.}.  Objects have to be saved using this stack around calls
that may cause a GC cycle to happen, and also while there is no
transaction running. In this simplified API, only
\lstinline!stm_allocate()!  and \lstinline!stm_commit_transaction()!
require saving object references.

The type \lstinline!object_t!  is special as it causes the
compiler~\footnote{Clang 3.5 with some patches to this address-space
  256 feature} to make all accesses through it relative to the $\%gs$
register.  With exceptions, nearly all accesses to objects managed by
the TM system should use this type so that the CPU will translate the
reference to the right version of the object.

@@

On startup, we reserve a big range of virtual memory with a call to
\lstinline!mmap()! and partition this space into $N+1$ segments. We
want to run $N$ threads in parallel while segment 0 is designated as
the \emph{sharing-segment }that is never assigned to a thread.

The next step involves using \lstinline!remap_file_pages()!, a Linux
system call, to establish the fully-shared configuration.  All pages
of segments $>0$ map to the pages of the sharing-segment.

However, the layout of a segment is not uniform and we actually
privatize a few areas again right away. These areas are illustrated in
Figure \ref{fig:Segment-Layout} and explained here:
\begin{description}[noitemsep]
\item [{NULL~page:}] This page is unmapped and will produce a
  segmentation violation when accessed. We use this to detect
  erroneous dereferencing of \lstinline!NULL! references.  All
  $\%gs{::}SO$ translated to linear addresses will point to NULL pages
  if SO is set to \lstinline!NULL!.
\item [{Segment-local~data:}] Some area private to the segment that
  contains segment-local information.
\item [{Read~markers:}] These are pages that store information about
  which objects were read in the current transaction running in this
  segment.
\item [{Nursery:}] This area contains all the freshly allocated
  objects (\emph{young objects}) of the current transaction. The GC
  uses pointer-bump allocation in this area to allocate objects in the
  first generation.
\item [{Old~object~space:}] These pages are the ones that are really
  shared between segments. They mostly contain old objects but also
  some young ones that were too big to allocate in the nursery.
\end{description}

\begin{figure*}[t]
  \begin{centering}
    \includegraphics[scale=0.8]{\string"segment layout\string".pdf}
    \par\end{centering}

@@

To improve this situation even more, we introduce the concept of
\emph{overflow objects}. If a minor collection needs to occur during a
transaction, we empty the nursery and mark each surviving object in
the old object space with an \lstinline!overflow_number!  globally
unique to the current transaction. That way we can still detect in a
medium-fast path inside barriers that the object still belongs to the
current transaction.

@@

As seen in the API (section~\ref{sub:Application-Programming-Interfac}),
we use a \emph{shadow stack} in order to provide precise garbage
collection.  Any time we call a function that possibly triggers a
collection, we need to save the objects that we need afterwards on the
shadow stack using \lstinline!STM_PUSH_ROOT()!.  That way, they will
not be freed. And in case they were young, we get their new location
in the old object space when getting them back from the stack using
\lstinline!STM_POP_ROOT()!.

@@

To add the object to the read set, for us it is enough to mark it as
read. Since this information needs to be local to the segment, we need
to store it in private pages. The area is called \emph{read markers
}and already mentioned in section \ref{sub:Setup}. This area can be
seen as a continuous array of bytes that is indexed from the start of
the segment by an object's reference ($SO$) divided by 16 (this
requires objects of at least 16 bytes in size). Instead of just
setting the byte to \lstinline!true!  if the corresponding object was
read, we set it to a \lstinline!read_version!  belonging to the
transaction, which will be incremented on each commit.  Thereby, we
can avoid resetting the bytes to \lstinline!false!  on commit and only
need to do this every 255 transactions. The whole code for the barrier
is easily optimizable for compilers as well as perfectly predictable
for CPUs:

\begin{lstlisting}[basicstyle={\footnotesize\ttfamily},tabsize=4]
void stm_read(SO):
    *(SO >> 4) = read_version
\end{lstlisting}

@@

The \textbf{fast path} of the write barrier is very simple. We only
need to check for the flag \lstinline!WRITE_BARRIER!  in the object's
header and call the slow path if it is set. This flag is set either if
the object is old and comes from an earlier transaction, or if there
was a minor collection which will add the flag again on all
objects. It is never set on freshly allocated objects that still
reside in the nursery.

\begin{lstlisting}[basicstyle={\footnotesize\ttfamily},tabsize=4]
void stm_write(SO):
	if SO->flags & WRITE_BARRIER:
		write_slowpath(SO)
\end{lstlisting}

@@

\begin{lstlisting}[basicstyle={\footnotesize\ttfamily},tabsize=4]
void write_slowpath(SO):
	// GC part:
	list_append(to_trace, SO)
	if is_overflow_obj(SO):
		SO->flags &= ~WRITE_BARRIER
		return
	// STM part
	stm_read(SO)
	lock_idx = SO >> 4
  retry:
	if write_locks[lock_idx] == our_num:
		// we already own it
	else if write_locks[lock_idx] == 0:
		if cmp_and_swap(&write_locks[lock_idx],
					    0, our_num):
			list_append(modified_old_objects, SO)
			privatize_pages(SO)
		else:
			goto retry
	else:
		w_w_contention_management()
		goto retry
	SO->flags &= ~WRITE_BARRIER
\end{lstlisting}

First comes the \emph{GC part}: In any case, the object will be added
to the list of objects that need tracing in the next minor collection
(\lstinline!to_trace!).  This is necessary in case we write a
reference to it that points to a young object. We then need to trace
it during the next minor collection in order to mark the young object
alive and to update its reference to the new location it gets moved
to. The check for \lstinline!is_overflow_obj()!  tells us if the
object was actually created in this transaction. In that case, we do
not need to execute the following \emph{TM part}.  We especially do
not need to privatize the page since no other transaction knows about
these ``old'' objects.

For TM, we first perform a read barrier on the object. We then try to
acquire its write lock. \lstinline!write_locks!  again is a simple
global array of bytes that is indexed with the SO of the object
divided by 16. If we already own the lock, we are done.  If someone
else owns the lock, we will do a write-write contention management
that will abort either us or the current owner of the object.  If we
succeed in acquiring the lock using an atomic
\lstinline!cmp_and_swap!, we need to add the object to the write set
(a simple list called \lstinline!modified_old_objects!)  and privatize
all pages belonging to it (copy-on-write).

In all cases, we remove the \lstinline!WRITE_BARRIER!  flag from the
object before we return. Thus, we never trigger the slow path again
before we do the next minor collection (also part of a commit) or we
start the next transaction.

@@

Aborting a transaction is rather easy. The first step is to reset the
nursery and all associated data structures. The second step is to go
over all objects in the write set (\lstinline!modified_old_objects!)
and reset any modifications in our private pages by copying from the
sharing-segment. What is left is to use \lstinline!longjmp()!  to jump
back to the location initialized by a \lstinline!setjmp()!  in
\lstinline!stm_start_transaction()!.  Increasing the
\lstinline!read_version!  is also done there.

@@

Committing a transaction needs a bit more work. First, we synchronize
all threads so that the committing one is the only one running and all
the others are waiting in a safe point. We then go through the write
set (\lstinline!modified_old_objects!)  and check the corresponding
\lstinline!read_markers!  in other threads/segments. If we detect a
read-write conflict, we do contention management to either abort us or
the other transaction, or to simply wait a bit.

After verifying that there are no conflicts anymore, we copy all our
changes done to the objects in the write set to all other segments,
including the sharing-segment. This is safe since we synchronized all
threads. We also need to push overflow objects generated by minor
collections to other segments, since they may reside partially in
private pages. At that point we also get a new
\lstinline!overflow_number!  by increasing a global one, so that it
stays globally unique for each transaction. Increasing the
\lstinline!read_version!  is then done at the start of a new
transaction.

@@

A requirement for performing a commit is to synchronize all threads so
that we can safely update objects in other segments. To make this
synchronization fast and cheap, we do not want to insert an additional
check regularly in order to see if synchronization is requested. We
use a trick relying on the fact that dynamic languages are usually
very high-level and thus allocate a lot of objects very regularly.
This is done through the function \lstinline!stm_allocate!  shown
below:

\begin{lstlisting}[basicstyle={\footnotesize\ttfamily},tabsize=4]
object_t *stm_allocate(ssize_t size_rounded):
    result = nursery_current
	nursery_current += size_rounded
	if nursery_current > nursery_end:
		return allocate_slowpath(size_rounded)
	return result
\end{lstlisting}

This code does simple pointer-bump allocation in the nursery. If there
is still space left in the nursery, we return
\lstinline!nursery_current!  and bump it up by
\lstinline!size_rounded!.  The interesting part is the check
\lstinline!nursery_current > nursery_end!  which will trigger the slow
path of the function to possibly perform a minor collection in order
to free up space in the nursery.

If we want to synchronize all threads, we can rely on this check being
performed regularly. So what we do is to set the
\lstinline!nursery_end!  to $0$ in all segments that we want to
synchronize. The mentioned check will then fail in those segments and
call the slow path. In \lstinline!allocate_slowpath!  they can simply
check for this condition and enter a safe point.

@@
